---
title: "Assignment: Staffing Validation"
author: "Dunja Novaković"
date: "2020-07-27"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Instructions

This assignment reviews the *Staffing Validation* analytical lecture. 
You will use the *staffing_validation.Rmd* file I reviewed in the video lectures to complete this assignment. 
You will *copy and paste* relevant code from that file and update it to answer the questions in this assignment. 
You will respond to questions in each section after executing relevant code to answer a question. 
You will submit this assignment to its *Submissions* folder on *D2L*.
You will submit this *(1)* completed **R Markdown** script and *(2)* a _HTML_ or _PDF_ rendered version of it to _D2L_ by the due date and time.
If you installed `TinyTeX` successfully, then I prefer a *PDF* version.

To start:

For any analytical project, you want to create a clear project directory structure.  
All materials from this course should exist in one folder on your computer.
Inside of that main course folder, you should create folders to store course documentation, lecture analytical projects, assignments analytical projects, etc. 
Inside of your folder for assignments analytical projects, you should create folder for this assignment named *staffing_validation*.

Any analytical project folder should contain inside it at least three additional folders named *scripts*, *data*, and *plots*.
Store this script in the *scripts* folder, the data for this assignment in the *data* folder, and any requested plots in the *plots* folder.
Each analytical project should also contain a **.Rproj** file in its top-level directory. 
Go to the *File* menu in *RStudio*, select *New Project...*, choose *Existing Directory*, go to the folder you created to contain this analytical project. 
Select it as the top-level directory for this **RStudio Project**.  

## Global Settings

The first code chunk sets the global settings for the remaining code chunks in the document.
Do *not* change anything in this code chunk.

```{r, setup, include=FALSE}
### specify echo setting for all code chunks
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

In this code chunk, we load packages we need for this assignment:

1. **here**,
2. **tidyverse**,
3. **skimr**,
4. **plotly**,
5. **broom**,
6. **modelr**,
7. **relaimpo**,
8. **interactions**, and
9. **jtools**.

We will use functions from these packages to import the data, examine the data, calculate summaries on the data, build regression models, and create visualizations from the data. 
Do *not* change anything in this code chunk.

```{r, libraries}
### load libaries for use in current working session
## here for workflow
library(here)

## tidyverse for data manipulation and plotting
# loads eight different libraries simultaneously
library(tidyverse)

## skimr for summary statistics
library(skimr)

## plotly for interactive 3D plots
library(plotly)

## broom to work with model objects
library(broom) 

## modelr to work with model objects
library(modelr)

## relaimp to calculate relative predictor importance
library(relaimpo)

## interactions to make 2D plots to capture interaction effects
library(interactions) 

## jtools for printing clean results tables
library(jtools)

## sandwich for simple slope analysis
library(sandwich)
```

## Task 1: Load Data

Load the data for this assignment.
We will use the same data as in the analytical lecture: **staff_sys_valid.csv**.

Use the appropriate functions to navigate to your *data* directory and import the database. 
Import the database as the object **staff_data**.
Use **skim_without_charts()** to calculate summaries on the data grouped by **race**.

**Question 1.1**: What is the mean **sjt** score for the **Hispanic** group?  

**Response 1.1**: *28.79*.

Compute the correlations among all of the numeric variables.

**Question 1.2**: Which two variables are most correlated?
Which two variables are least correlated?

**Response 1.2**: *Most correlated: _sjt_ and _work_samp_. Least correlated: _job_perf_ and _cust_serv_*.

```{r, task1}
##### Q1.1
#### import database
### use here() to locate file in our project directory;
### use read_csv() to import the data
                       ## file path    
staff_data <- read_csv(here("data", "staff_sys_valid.csv"), 
  ## column types;
  ## treat gender and race as factor variables
  col_types = cols(gender = col_factor(), race = col_factor()))

### distribution summaries
## call data
staff_data %>%
  ## group by factors
  group_by(race) %>%
  ## skim
  skim_without_charts()

##### Q1.2
### correlations
## call data
staff_data %>%
  ## select
  select_if(is.numeric) %>%
  ## correlations
  cor() 
```

## Task 2: Simple OLS Regression

Estimate a simple OLS regression model where you predict **cust_serv** scores from **sjt**.
Name the model **mod_1**.
Examine the results with the **summary()** function.

**Question 2.1**: How do you correctly interpret the **sjt** regression coefficient?

**Response 2.1**: *The sjt regression coefficient tells us the expected increase in cust_serv for one unit increase in sjt. For one unit increase in sjt, we expect a 0.477599 increase in cust_serv*.

Produce a scatterplot with **sjt** on x-axis and **cust_serv** on y-axis.
Fit a regression line through the data points.

**Question 2.2**: Do individuals with a **sjt** score of 35 or 25 have higher **cust_serv** scores?

**Response 2.2**: *Individuals with a sjt score of 35 have higher cust_serv scores*.

Apply **augment()** to **mod_1** and save the result as **mod_1_fit**.
Open the spreadsheet view of **mod_1_fit**.

**Question 2.3**: What is the third person's actual **cust_serv** score and this person's predicted **cust_serv** score from the model?

**Response 2.3**: *Actual score: 20. Predicted score: 25.32529*.

Calculate all of the unique predicted values from **mod_1** for each unique value of **sjt**.

**Question 2.4**: What is the predicted value on **cust_serv** for someone with a **sjt** score of *25*?

**Response 2.4**: *23.41490*.

```{r, task2}
##### Q2.1
#### simple OLS regression model
### build model
mod_1 <- lm(cust_serv ~ sjt, data = staff_data)

## summary of results
# fuller output
summary(mod_1)

##### Q2.2
### plot data and prediction line
## call data and set mapping
ggplot(staff_data, aes(x = sjt, y = cust_serv)) +
  ## jitter geometry
  geom_jitter(width = 0.4, height = 0.2, alpha = 0.5) +
  ## smooth geometry
  geom_smooth(method = "lm")

##### Q2.3
## compute fitted values for all individuals in the sample
mod_1_fit <- augment(mod_1)

##### Q2.4
## examine unique fitted values
# call data
mod_1_fit %>%
  # call predictor
  data_grid(sjt) %>%
  # calculate unique fitted values
  add_predictions(mod_1)
```

## Task 3: Multiple OLS Regression

Estimate a multiple OLS regression model where you predict **cust_serv** scores from **emot_intel** and **str_int** scores.
Name the model **mod_2**.
Examine the results with the **summary()** function.

**Question 3.1**: How do you correctly interpret the **str_int** regression coefficient?
What is the *multiple R-squared* for the model?
On average, how wrong are the model's predictions (i.e., look at the *residual standard error*) in **cust_serv** units?

**Response 3.1**: *The str_int regression coefficient tells us the expected increase in cust_serv for one unit increase in str_int. For one unit increase in str_int, we expect a 0.42642 increase in cust_serv. The multiple R-squared for this model is equal to 0.2922. On average, the model's predictions are wrong by 3.487*.

Apply **augment()** to **mod_2** and save the result as **mod_2_fit**.
Open the spreadsheet view of **mod_2_fit**.

**Question 3.2**: What is the seventh person's actual **cust_serv** score and this person's predicted **cust_serv** score from the model?

**Response 3.2**: *Actual score: 24. Predicted score: 25.70824*.

Calculate all of the unique predicted values from **mod_2** for each unique combination of **emot_intel** and **str_int**.

**Question 3.3**: What is the predicted value on **cust_serv** for someone with a **emot_intel** score of *12* and **str_int** score of *21*?

**Response 3.3**: *18.53118*.

Produce a 3D scatterplot using **plotly** functions.
Place **emot_intel** on the x-axis, **str_int** on the y-axis, and **cust_serv** on the z-axis.
Label the axes using *"Emot. Intel."*, *"Str. Int."*, and *"Cust. Serv."*, respectively.
Fit a regression plane through the data points adjusting the relevant code from the analytical lecture.
Make sure to compute **x_vals** for **emot_intel** and **y_vals** for **str_int**.
Make sure to name **pred_data** columsn with *"emot_intel"* first and *"str_int"* second.
Change the name of the colorbar to *"Cust. Serv."*.
Print the plot after creating it.
Click the **Zoom** button to pop out the plot.
Interact with the plot by rotating it and zooming in and out.

**Question 3.4**: Looking at the regression plane, do individuals with scores in the *upper-30s* or *mid-20s* on **emot_intel** and **str_int** have higher **cust_serv** scores?

**Response 3.4**: *Individuals with scores in the upper-30s on emot_intel and str_int have higher cust_serv_scores*.

```{r, task3}
##### Q3.1
#### multiple OLS regression model
### build model
mod_2 <- lm(cust_serv ~ emot_intel + str_int, data = staff_data)

## summary of results
# fuller output
summary(mod_2)

##### Q3.2
## compute fitted values for all individuals in the sample
mod_2_fit <- augment(mod_2)

##### Q3.3
## examine unique fitted values
# call data
mod_2_fit %>%
  # call predictor
  data_grid(emot_intel, str_int) %>%
  # calculate unique fitted values
  add_predictions(mod_2) 

##### Q3.4
### 3D plot without regression plane
## select data and assign variables
scatter3d <- plot_ly(data = mod_2_fit, 
                     x = ~emot_intel, 
                     y = ~str_int, 
                     z = ~cust_serv, 
                     opacity = 0.6) %>% 
  ## add data points  
  add_markers() %>% 
  ## add axes labels
  layout(scene = list(
          xaxis = list(title = "Emot. Intel."),
          yaxis = list(title = "Str. Int."),
          zaxis = list(title = "Cust. Serv.")))

### plot regression plane
## set-up predictor value grid
# x-axis unique values
x_vals <- seq(min(mod_2_fit$emot_intel), max(mod_2_fit$emot_intel), length.out = 50)

# y-axis unique values
y_vals <- seq(min(mod_2_fit$str_int), max(mod_2_fit$str_int), length.out = 50)

# grid of predictor values
pred_data <- expand.grid(x_vals, y_vals)

# name the columns of the prediction grid
names(pred_data) <- c("emot_intel", "str_int")

## use regression model to generate prediction values
fit_vals <- predict(mod_2, newdata = pred_data)

# organize prediction values in a matrix
plane <- matrix(fit_vals, nrow = 50, ncol = 50, byrow = T)

## call the 3D scatterplot
scatter3d_plane <- scatter3d %>% 
  ## add regression plane to 3D scatterplot
              # x-axis values
  add_surface(x = ~ x_vals, 
              # y-axis values
              y = ~ y_vals, 
              # z-axis values for plane
              z = ~ plane, 
              # include z-axis color scale
              showscale = T) %>%
  ## add title for color scale
  colorbar(title = "Cust. Serv.")
## print plot
scatter3d_plane
```

## Task 4: Staffing System

Estimate a multiple OLS regression model where you predict **cust_serv** scores from:

1. **proactive**, 
2. **emot_intel**, 
3. **sjt**, 
4. **work_samp**, and 
5. **str_int** scores scores.

Name the model **mod_3**.
Examine the results with the **summary()** function.

**Question 4.1**: How do you correctly interpret the **work_samp** regression coefficient?
What is the *multiple R-squared* for the model?
On average, how wrong are the model's predictions (i.e., look at the *residual standard error*) in **cust_serv** units?

**Response 4.1**: *The work_samp regression coefficient tells us the expected increase in cust_serv for one unit increase in work_samp. For one unit increase in work_samp, we expect a 0.20141 increase in cust_serv. The multiple R-squared for the model is equal to 0.3396. On average, the model's predictions are wrong by 3.368. *.

Apply **augment()** to **mod_3** and save the result as **mod_3_fit**.
Open the spreadsheet view of **mod_3_fit**.

**Question 4.2**: What is the twenty-second person's actual **cust_serv** score and this person's predicted **cust_serv** score from the model?

**Response 4.2**: *Actual score: 31. Predicted score: 24.95111*.

Use **mod_3_fit** to print the actual criterion (i.e., **cust_serv**) rank order and the model (i.e., **.fitted**) rank order of individuals.

**Question 4.3**: Which individual has the highest observed **cust_serv** score?
Which individual does the model predict to have the highest **cust_serv** score?
Out of the *top 10* observed **cust_serv** scores, how many did the model predict to be in the *top 10* of **cust_serv** scores?

**Response 4.3**: *The individual with the id equal to 3751 has the highest observed cust_serv score. The model predicts the individual with the id equal to 5051 to have the highest cust_serv score. The model predicted 2 out of the top 10 observed cust_serv scores*.

Evaluate staffing decisions when the hiring threshold is *27* and the criterion threshold is *26*.
Plot the two thresholds with fitted values on the x-axis and criterion values on the y-axis.
Compute the number of true positives, true negatives, false positives, and false negatives and save it to an object named **mod_3_acc**.
Calculate the overall, positive, negative, sensitivity, and specificity accuracy of staffing decisions with this model and these thresholds.

**Question 4.4**: How many total *false positive* decisions would be made?
How many total *true negative* decisions would be made?
Which accuracy measure would be highest?
Which accuracy measure would be lowest?

**Response 4.4**: *Total false positives: 469. Total true negatives: 4817. The highest accuracy measure is specificity. The lowest accuracy measure is sensitivity*.

Compute the relative importance of predictors in **mod_3**.

**Question 4.5**: How much of the total *R-squared* is attributed to **emot_intel**?
Which predictor accounts for the most *R-squared*?

**Response 4.5**: *Approx. 6.16% of the total R-squared is attributed to emot_intel. The predictor str_int accounts for the most R-squared*.

```{r, task4}
##### Q4.1
#### multiple OLS regression model
### build model
mod_3 <- lm(cust_serv ~ proactive + emot_intel + sjt + work_samp + str_int, 
            data = staff_data)

## summary of results
# fuller output
summary(mod_3)

##### Q4.2
## compute fitted values for all individuals in the sample
mod_3_fit <- augment(mod_3)

##### Q4.3
### actual criterion rank order
## call data
mod_3_fit %>% 
  ## add id
  mutate(id = 1:nrow(mod_3_fit)) %>%
  ## select variables
  dplyr::select(id, cust_serv, .fitted) %>%
  ## arrange by criterion values
  arrange(desc(cust_serv)) 

### model rank order
## call data
mod_3_fit %>% 
  ## add id
  mutate(id = 1:nrow(mod_3_fit)) %>%
  ## reorder variables
  dplyr::select(id, cust_serv, .fitted) %>%
  ## arrange by fitted values
  arrange(desc(.fitted)) 

##### Q4.4
### plot data and prediction line
## thresholds
# hiring
hire_thresh <- 27

# criterion
crit_thresh <- 26

## call data and set mapping
ggplot(mod_3_fit, aes(x = .fitted, y = cust_serv)) +
  ## jitter geometry
  geom_jitter(width = 0.4, height = 0.2, alpha = 0.5) +
  ## criterion value threshold for successful customer service
  geom_hline(yintercept = crit_thresh, color = "red", size = 2) +
  ## predicted value threshold for hiring
  geom_vline(xintercept = hire_thresh, color = "green", size = 2)

### evaluate accuracy of predictions
## name result and choose data
mod_3_acc <- mod_3_fit %>%
  ## summarize
            # true positives
  summarize(tp = sum(.fitted >= hire_thresh & cust_serv >= crit_thresh),
            # true negatives
            tn = sum(.fitted < hire_thresh & cust_serv < crit_thresh),
            # false positives
            fp = sum(.fitted >= hire_thresh & cust_serv < crit_thresh),
            # false negatives
            fn = sum(.fitted < hire_thresh & cust_serv >= crit_thresh))

## accuracy computations
mod_3_acc %>%
  # overall accuracy
  summarize(overall = (tp + tn)/(tp + tn + fp + fn),
            # positive accuracy
            positive = tp/(tp + fp),
            # negative accuracy
            negative = tn/(tn + fn),
            # sensitivity
            sensitivity = tp/(tp + fn),
            # specificity
            specificity = tn/(tn + fp))

##### Q4.5
### predictor importance
## specify model and type of relative importance
calc.relimp(mod_3, type = "car")


```

## Task 5: Evaluate Groups

Add mean-centered **emot_intel** to **staff_data**.
Name the new variable **emot_intel_cent**.
Examine the contrasts of **gender**.
Estimate a model where **cust_serv** scores are predicted by **emot_intel_cent** and **gender** scores including the *interaction* between the two predictors. 
Name the model **mod_4**.
Examine the results with the **summary()** function.

**Question 5.1**: What is the regression coefficient for the interaction effect of the two predictors?

**Response 5.1**: *-0.13593*.

Produce the scatterplot and the regression lines for each **gender**.
Calculate the simple slopes for each **gender** from the model.

**Question 5.2**: Is the difference between *female* and *male* predicted **cust_serv** scores greater when individuals are *10* units *above* or *below* the mean on **emot_intel**? 
How do you interpret the simple slope for *females*?

**Response 5.2**: *The difference between female and male predicted cust_serv scores is greater when individuals are 10 units above the mean on emot_intel. The estimated slope for females is equal to 0.45 and is higher than the slope for males. It indicates a stronger connection between emotional intelligence and customer service scores for females*.

Examine the contrasts of **race**.
Estimate a model where **cust_serv** scores are predicted by **emot_intel_cent** and **race** scores including the *interaction* between the two predictors. 
Name the model **mod_5**.
Examine the results with the **summary()** function.

**Question 5.3**: What is the regression coefficient for the interaction effect of **emot_intel_cent** and the *Hispanic* group?

**Response 5.3**: *0.07343*.

Produce the scatterplot and the regression lines for each **race**.
Calculate the simple slopes for each **race** from the model.

**Question 5.4**: Which group is predicted to have the lowest **cust_serv** scores regardless of **emot_intel_cent** scores?
Which group has the second-strongest relationship between **emot_intel_cent** and **cust_serv** (i.e., which group has the second-highest simple slope)?

**Response 5.4**: *White people are predicted to have the lowest cust_serv scores. Black people the second-strongest relationship between emot_intel_cent and cust_serv*.

```{r, task5}
##### Q5.1
#### data preparation
### center continuous predictor
## call data
staff_data <- staff_data %>%
  ## mutate
  mutate(emot_intel_cent = emot_intel - mean(emot_intel))

## examine contrasts of categorical predictor
# gender
contrasts(staff_data$gender)

#### multiple OLS regression model
### build model
mod_4 <- lm(cust_serv ~ emot_intel_cent * gender, data = staff_data)

## summary of results
# fuller output
summary(mod_4)

##### Q5.2
### plot
## call data
ggplot(staff_data, aes(x = emot_intel_cent, y = cust_serv, color = gender)) +
  ## jitter geometry
  geom_jitter(width = 0.4, height = 0.2, alpha = 0.5) +
  ## smooth geometry
  geom_smooth(aes(linetype = gender), color = "black", method = "lm")

### calculate simple slopes
## specify model
sim_slopes(model=mod_4, 
  ## specify x-axis variable
  pred = emot_intel_cent, 
  ## specify moderator variable
  modx = gender)

##### Q5.3
### build model
## contrasts for race
contrasts(staff_data$race)

## model
mod_5 <- lm(cust_serv ~ emot_intel_cent * race, data = staff_data)

## summary of results
# fuller output
summary(mod_5)

##### Q5.4
### plot
## call data
ggplot(staff_data, aes(x = emot_intel_cent, y = cust_serv, color = race)) +
  ## jitter geometry
  geom_jitter(width = 0.4, height = 0.2, alpha = 0.5) +
  ## smooth geometry
  geom_smooth(aes(linetype = race), color = "black", method = "lm")

### calculate simple slopes
## specify model
sim_slopes(mod_5, 
  ## specify x-axis variable
  pred = emot_intel_cent, 
  ## specify moderator variable
  modx = race)

```