---
title: "Assignment: Flexdashboards"
author: "Dunja Novaković"
output:
  flexdashboard::flex_dashboard:
    source_code: embed
    orientation: columns
    theme: paper
---

```{r, libraries, include=FALSE}
### tidyverse for data wrangling and visualization
library(tidyverse)

### here to navigate project directory
library(here)

### flexdashboard to build dashboards in R Markdown
library(flexdashboard)

### plotly for web-friendly, interactive plots
library(plotly)

### DT for web-friendly, interactive tables
library(DT)

### rpart for decision trees
library(rpart)
```

```{r, data, include=FALSE}
### import a space delimited data file
## import data
emp_data <- read_delim(here("data", "emp.dat"), delim = " ",
                       ## column types
                       col_types = "dfffffddddd")

### sample data
emp_samp <- emp_data %>%
  ## sample
  sample_n(1500)

## number without work accident
num_wout_work_acc <- emp_samp %>%
  ## filter for employees who haven't experienced work accidents
  filter(work_accident == "No") %>%
  ## count
  nrow()

## number with work accident
num_with_work_acc <- emp_samp %>%
  ## filter for employees who have experienced a work accident
  filter(work_accident == "Yes") %>%
  ## count
  nrow()

## training data
emp_train <- emp_samp %>%
  ## sample a fraction
  sample_frac(0.65)

## testing data
emp_test <- emp_samp %>%
  ## find the difference between data
  setdiff(emp_train)
```

Plots
================

Column {.tabset}
-----------------------------------------------------------------------

### Performance v. Tenure

```{r, tenure}
### plot data
## scatterplot for performance v. tenure
# call data and set aesthetics
ggplot(emp_samp, aes(x=tenure, y = last_evaluation)) +
  # point geometry
  geom_point(alpha = 0.5) +
  # smooth geometry
  geom_smooth(method = "lm") +
  # aesthetic labels
  labs(x = "Tenure", y = "Performance") 
```

### Performance v. Number of Projects

```{r, projNum}
### plot data
## scatterplot for performance v. number_project
# call data and set aesthetics
ggplot(emp_samp, aes(x=number_project, y = last_evaluation)) +
  # point geometry
  geom_point(alpha = 0.5) +
  # smooth geometry
  geom_smooth(method = "lm") +
  # aesthetic labels
  labs(x = "Number of Projects", y = "Performance") 
```

### Performance v. Average Hours per Month

```{r, avgMtHr}
### plot data
## scatterplot for performance v. avg_month_hours
# call data and set aesthetics
ggplot(emp_samp, aes(x=avg_month_hours, y = last_evaluation)) +
  # point geometry
  geom_point(alpha = 0.5) +
  # smooth geometry
  geom_smooth(method = "lm") +
  # aesthetic labels
  labs(x = "Average Hours per Month", y = "Performance") 
```

### Performance v. Job Satisfaction

```{r, jobSat}
### plot data
## scatterplot for performance v. job_sat
# call data and set aesthetics
ggplot(emp_samp, aes(x=job_sat, y = last_evaluation)) +
  # point geometry
  geom_point(alpha = 0.5) +
  # smooth geometry
  geom_smooth(method = "lm") +
  # aesthetic labels
  labs(x = "Job Satisfaction", y = "Performance") 
```

Low Salary {data-navmenu=Tables} 
================

Column
-----------------------------------------------------------------------

### Low Salary

```{r, salLow}
## call data
emp_samp %>%
  ## filter
  filter(salary == "low") %>%
  ## select variables
  select(emp_id, tenure, number_project, avg_month_hours, job_sat, last_evaluation) %>%
  ## datatable
                # remove row names
  DT::datatable(rownames = FALSE, 
                # add buttons
                extensions = "Scroller",
                # options
                options = list(
                  # table, length, pagination, scrolling
                  dom = "tlpS",
                  # scroller
                  scrollY = "200px"
                ))
```

Medium Salary {data-navmenu=Tables} 
================

Column
-----------------------------------------------------------------------

### Medium Salary

```{r, salMed}
## call data
emp_samp %>%
  ## filter
  filter(salary == "medium") %>%
  ## select variables
  select(emp_id, tenure, number_project, avg_month_hours, job_sat, last_evaluation) %>%
  ## datatable
                # remove row names
  DT::datatable(rownames = FALSE,
                # add buttons
                extensions = "Scroller",
                # options
                options = list(
                  # table, length, pagination, scrolling
                  dom = "tlpS",
                  # scroller
                  scrollY = "200px"
                ))
```

High Salary {data-navmenu=Tables} 
================

Column
-----------------------------------------------------------------------

### High Salary

```{r, salHigh}
## call data
emp_samp %>%
  ## filter
  filter(salary == "high") %>%
  ## select variables
  select(emp_id, tenure, number_project, avg_month_hours, job_sat, last_evaluation) %>%
  ## datatable
                # remove row names
  DT::datatable(rownames = FALSE,
                # add buttons
                extensions = "Scroller",
                # options
                options = list(
                  # table, length, pagination, scrolling
                  dom = "tlpS",
                  # scroller
                  scrollY = "200px"
                ))
```


Gauges 
================
Column
-----------------------------------------------------------------------

### Number of Employees Without a Work Accident

```{r, numWoutWA}
## value box
         # value
valueBox(num_wout_work_acc,
         # caption
         caption = "Number of Employees Without a Work Accident",
         # icon
         icon = "fa-chart-bar")
```

### Percent of Employees Without a Work Accident

```{r, propWoutWA}
## gauge for employees without a work accident
      # value
gauge(value = round(100*num_wout_work_acc/nrow(emp_samp), 2),
      # minimum and maximum
      min = 0, max = 100,
      # sectors
      sectors = gaugeSectors(
        success = c(90, 100),
        warning = c(50, 89),
        danger = c(0, 49)
      ),
      # percentage calculation
      symbol = "%")
```


Column
-----------------------------------------------------------------------

### Number of Employees With a Work Accident

```{r, numWithWA}
## value box
         # value
valueBox(num_with_work_acc,
         # caption
         caption = "Number of Employees With a Work Accident",
         # icon
         icon = "fa-chart-bar")
```

### Percent of Employees With a Work Accident

```{r, propWithWA}
## gauge for employees with a work accident
      # value
gauge(value = round(100*num_with_work_acc/nrow(emp_samp), 2),
      # minimum and maximum
      min = 0, max = 100,
      # sectors
      sectors = gaugeSectors(
        success = c(90, 100),
        warning = c(50, 89),
        danger = c(0, 49)
      ),
      # percentage calculation
      symbol = "%")
```


Predictions {.storyboard}
================

### Evalute Accuracy of a Decision Tree for Turnover Predictions

```{r, turnover}
## model
mod_1_train <- rpart(left ~ . - emp_id, data = emp_train)

## class predictions
mod_1_test_class <- predict(mod_1_train, newdata = emp_test, 
                            type = "class")

## tibble
tibble(actual = emp_test$left, pred = mod_1_test_class) %>%
  ## group by actual values on outcome
  count(actual, pred) %>% 
  ## create accuracy type
  mutate(type = c("tp", "fn", "fp", "tn")) %>% 
  ## pivot wider
  pivot_wider(id_cols = -c("actual", "pred"), 
              names_from = "type", values_from = "n") %>% 
  ## summarize accuracy
  summarize(Positive = round(tp/(tp + fp), 2), 
            Negative = round(tn/(tn + fn), 2), 
            Sensitivity = round(tp/(tp + fn), 2), 
            Specificity = round(tn/(tn + fp), 2)) %>% 
  ## pivot longer
  pivot_longer(cols = everything(),
               names_to = "acc", values_to = "pct")  %>%
  ## mutate to factor
  mutate(acc = as_factor(acc),
         acc = fct_reorder(acc, pct)) %>%
  ## plot
  ggplot(aes(x = acc, y = pct, fill = acc)) +
    ## bar geometry
    geom_col() +
    ## text geometry
    geom_text(aes(label = paste0(pct*100, "%")), size = 3,
            position = position_stack(vjust = 0.5), color = "white") +
    ## y-axis
    scale_y_continuous(labels = scales::percent_format()) +
    ## aesthetic labels
    labs(x = "Accuracy Type", y = "Percentage", fill = "Predicted") +
    ## flip coordinates
    coord_flip() +
    ## theme
    theme(legend.position = "none")
```

***

This highlights the accuracy of a single decision tree to predict turnover on test data from the available predictors.

### Evaluate Accuracy of OLS Regression for Performance Predictions

```{r, performance}
## model
mod_2_train <- lm(last_evaluation ~ . - emp_id, data = emp_train)

# regression predictions
mod_2_test_pred <- predict(mod_2_train, newdata = emp_test)

## tibble
tibble(actual = emp_test$last_evaluation, pred = mod_2_test_pred) %>%
  ## plot
  ggplot(aes(x = pred, y = actual)) +
    ## point geometry
    geom_point() +
    ## smooth geometry
    geom_smooth(method = "lm", formula = "y ~ x") +
    ## aesthetic labels
    labs(x = "Predicted Performance", y = "Actual Performance")
```

***

This highlights the accuracy of OLS regression to predict performance on test data from the available predictors.
